Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-12 paper-new/concrete-relational-model.tex
more comments
updates to relational model section: discuss interpretation of syntax and revise section on adequacy

We also enhance our error domains with ordering and bisimilarity
relation, and we impose a condition relating the $\theta$ morphism to
the bisimilarity relation. More concretely:
%
\begin{definition}
An \textbf{error domain} $B$ consists of a predomain $B$ along with the following data:
\begin{itemize}
    \item A distinguished ``error'' element $\mho_B \in B$
    \item A morphism of predomains $\theta_B \colon \later B \to B$
    \item For all $x : B$ we have $\theta_B(\nxt\, x) \bisim_B x$
\end{itemize}\max{Don't we also need $\forall x. \mho \le x$?}
\end{definition}
%
For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B
\circ \nxt$.
%
A \textbf{morphism of error domains} is simply a morphism of the underlying
predomains that preserves the error element and the $\theta$ map.

@@

Next we define an analogous notion of \emph{relation} between error domains $B$
and $B'$. An error domain relation $d : B \rel B'$ is a relation on the
underlying posets that preserves error and respects $\theta$. More formally:
%
\begin{definition}
  Let $B$ and $B'$ be error domains. An \textbf{error domain relation} between
  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
  \begin{enumerate}
     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.\max{can we please use infix notation consistently for relations?}
     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t ((\tilde{x}_t, \tilde{y}_t) \in
     d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in d$. 
  \end{enumerate}
\end{definition}
%
As with predomain relations, we can compose error domain
relations. However, because of a technical restriction\max{technical
  restriction?? isn't is simpler/correct to say that the composition
  of underlying relations is not an error domain relation?} involving
the interaction between propositional truncation and the later
modality, we cannot define the composition of error domain relations
as the composition of their underlying predomain relations. Instead,
we define composition of error domain relations $d$ on $B_1$ and $B_2$
and $d'$ on $B_2$ and $B_3$ inductively to be the least relation
containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
upward-closed under $\le_{B_3}$, respects error, and preserves
$\theta$.  Specifically, it is defined inductively by the following
rules:\max{what rules??}
% %
% \begin{mathpar}
%     \inferrule*[right = Comp]
%     {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
%     {b_1 \mathbin{d \relcomp d'} b_3}

%     \inferrule*[right = DnClosed]
%     {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
%     {b_1' \mathbin{d \relcomp d'} b_3}

%     \inferrule*[right = UpClosed]
%     {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
%     {b_1 \mathbin{d \relcomp d'} b_3'}

%     \inferrule*[right = PresErr]
%     { }
%     {\mho_{B_1} \mathbin{d \relcomp d'} b_3}

%     \inferrule*[right = PresTheta]
%     {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
%     {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
% \end{mathpar}
% %

@@

The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
act on predomains and error domains and also have actions on their morphisms,
relations, and squares. The functor $U$ simply takes the underlying predomain
relation of an error domain relation. Given a relation $c_1 : A_1 \rel A_1'$ and
$c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2) \rel
(A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$ if
and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error\max{preserves error?}
and respects $\theta$ follows from the corresponding properties for $d$.

@@

To accomplish this, we equip every predomain $A$ with a monoid $M_A$ of
\emph{syntactic perturbations}, as well as a means of \emph{interpreting} these
syntactic perturbations as \emph{semantic} perturbations on $A$, i.e., as
endomorphisms that are bisimilar to the identity.
% Moreover, we want the
% resulting endomorphisms to be extensionally equivalent to the identity function,
% a condition we express by saying that they are weakly bisimilar to the identity
% morphism on $A$. 
More formally, for every $A$ we require a homomorphism of monoids $i_A : M_A \to
\morbisimid{A} := \{ f : A \to A \mid f \bisim \id_A \}$. We likewise equip
every error domain $B$ with a monoid $M_B$ and interpretation homomorphism $i_B
: M_B \to \morbisimid{B}$. This grouping of data deserves its own definition: We
call the triple $(A, M_A, iA)$ a \textbf{value object} as it will be the
denotation of value types in our final model. Likewise, the triple $(B, M_B,
i_B)$ will be called a \textbf{computation object}.\max{make this a proper \textit{Definition} and define morphisms as well}

\max{idea: move this after the quasi-representability part and say that this structure is what's needed to make them compose}
We must impose an additional condition on predomain relations $c$ and error
domain relations $d$ that specify how they interact with these perturbations. In
particular, we need to be able to ``push'' and ``pull'' perturbations along
relations $c$ and $d$. The intuition for this requirement comes from the
construction of the square corresponding to the original UpL rule from the
square for the simplified version as was shown in Section
\ref{sec:towards-relational-model}. Recall that we horizontally composed the
simplified UpL square for $c$ with the identity square for $c'$ on the right.
Now that the square for UpL involves a perturbation on the right instead of the
identity, this construction will not work unless we can ``push'' the
perturbation on $A_2$ to one on $A_3$. That is, given a relation $c : A \rel A'$
and a syntactic perturbation $m_A \in M_A$, we need to be able to turn it into a
syntactic perturbation $m_{A'} \in M_{A'}$ such that the resulting semantic
perturbations obtained by applying the respective homomorphisms $i_A$ and
$i_{A'}$ form a square. Dually, we will need to ``pull'' a perturbation from
right to left. This notion is made formal by the following definition:
%
\begin{definition}
    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects,
    and let $c : A \rel A'$ be a predomain relation. A
    \textbf{push-pull structure} for $c$ (with respect to $M_A$ and
    $M_{A'}$) consists of monoid homomorphisms $\push : M_A \to M_A'$
    and $\pull : M_A' \to M_A$ such that for all $m_A \in M_A$ there
    is a square $i_A(m_A) \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all
    $m_{A'} \in M_{A'}$ there is a square $i_A(\pull\, m_{A'})
    \ltsq{c}{c} i_{A'}(m_{A'})$. A push-pull structure for error
    domain relations is defined similarly.
\end{definition}

With the notion of syntactic perturbation defined, we can now make formal the
final aspect of the model: the notion of quasi-represntability of relations.
%
\begin{definition}[quasi-left-representable relations]\label{def:quasi-left-representable}
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: (1) $\upl$: $e \ltsq{c}{\le_{A'}} i_{A'}(\delre_c)$, and
(2) $\upr$: $i_A(\delle_c) \ltsq{\le_{A}}{c} e$.
\end{definition}
%
Observe that this weakens the previous notion of representability, since under
that definition the perturbations were required to be the identity.
%
We make the analogous definition of quasi-left-representability for error domain
relations $d$.

@@

\max{see above comment about reordering these concepts}
do not in general compose: in order to construct
the needed squares for the composition of two relations, we require that the
relations have push-pull structures.

@@

It remains to specify the actions of the functors on value and computation
objects and relations.
% Because we have augmented the definitions of object and relation for our model,
% we need to specify the actions of the functors $F$, $U$, $\times$, and $\arr$ on
% these objects and relations. 
% For instance, the functor $F$ now acts on \emph{value objects} $(A, M_A, i_A)$,
% which means we need to specify the monoid $M_{FA}$ and interpretation $i_{FA}$
% corresponding to the error domain $FA$.

\subsubsection{Constructions involving Perturbations}\label{sec:perturbation-constructions}
\max{this section needs prose}
We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
i_A)$ we define the syntactic perturbations $M_{\li A}$ to be $\mathbb{N} \oplus
M_A$, where $\oplus$ denotes the free product of monoids (the coproduct in the
category of monoids). The intuition behind needing to include $\mathbb{N}$ comes
from the example of the downcast for $\iarr$. Recall that we adjusted the
squares corresponding to the DnL and DnR rules by adding $(\delta \circ
\eta)^\dagger$ on the side opposite the downcast. Thus, we need a syntactic
perturbation in $M_{\li A}$ that will be interpreted as $(\delta \circ
\eta)^\dagger$. We accomplish this by taking the coproduct with $\mathbb{N}$.
%
Then to interpret the perturbations as endomorphisms, by the universal property
of the coproduct of monoids it suffices to define two homomorphisms, one from
$\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
The first homomorphism sends the generator $1 \in \mathbb{N}$ to the error
domain endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in
M_A$ to $\li(i_A(m_A))$ and we observe in both cases that the resulting
endomorphisms are bisimilar to the identity since the action of $\li$ on
morphisms preserves bisimilarity.

@@

\subsection{The Dynamic Type}
We now describe the denotation of the dynamic type. Analogously to how we
defined the dynamic type in Section \ref{sec:concrete-term-model}, the predomain
representing the dynamic type is defined using a combination of least fixpoint
and guarded recursion as the solution to the equation of \emph{predomains}
%
\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr \li D). \]
%
As before, we name the constructors $\inat$, $\itimes$, and $\iarr$
respectively, and the upcasts $e_{\nat}$, $e_\times$, and $e_\to$ are the same
as the constructors except for $e_{\iarr}$ which includes a $\nxt$.

%
% We define $e_\mathbb{N} : \mathbb{N} \to D$ to be the $\text{nat}$ constructor,
% $e_\times : D \times D \to D$ to be $\text{times}$, and $e_\to : U(D \arr F D)$
% to be the morphism $\nxt$ followed by $\text{fun}$.

The ordering $\le_D$ is given by:\max{this is kind of obvious IMO maybe it can go in the appendix}
%
\begin{align*}
    \inat(n) \le \inat(n') 
        &\iff n = n' \\
    \itimes (d_1, d_2) \le \itimes (d_1', d_2')
        &\iff d_1 \le_D d_2 \text{ and } d_1' \le_D d_2'\\
    \iarr(\tilde{f}) \le \iarr(\tilde{f'}) 
        &\iff \later_t(\tilde{f}_t \le \tilde{f'}_t),
\end{align*}
%
and the bisimilarity relation is defined analogously.
%
We define three relations involving $D$ as follows. We define $\inat :
\mathbb{N} \rel D$ by $(n, d) \in \inat$ iff $e_\nat(n) \le_D d$. We similarly
define $\itimes : D \times D \rel D$ by $((d_1, d_2), d) \in \itimes$ iff
$e_\times(d_1, d_2) \le_D d$, and we define $\text{inj}_\to : U(D \arr \li D)
\rel D$ by $(f, d) \in \iarr$ iff $e_\to(f) \le_D d$.

Now we define the perturbations for $D$. Recall that to each predomain $A$ we
associate a monoid $M_A$ of perturbations and a homomorphism into the monoid of
endomorphisms bisimilar to the identity, and likewise for error domains. It may
seem as though we need to define the perturbations for $D$ by guarded recursion,
but in fact we define them via least-fixpoint in the category of monoids:
%
\( M_D \cong (M_{D \times D}) \oplus M_{U(D \to \li D)}. \)
%
We now explain how to interpret these perturbations as endomorphisms.

@@

One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
\bisim \id \}$. We claim\max{why we claim? Just say it's true!} that the three relations $\inat$, $\itimes$, and
$\iarr$ satisfy the push-pull property. As an illustrative case, we establish
the push-pull property for the relation $\iarr$. We define $\pull_{\iarr} : M_D
\to M_{U(D \arr \li D)}$ by giving a homomorphism from $M_{D \times D} \to
M_{U(D \arr \li D)}$ and from $M_{U(D \to \li D)} \to M_{U(D \arr \li D)}$. The
former is the trivial homomorphism sending everything to the identity element,
while the latter is the identity homomorphism.
%
Conversely, we define $\push_{\iarr} : M_{U(D \arr \li D)} \to M_D$ as the
inclusion into the coproduct.
%
Showing that the relevant squares exist involving push and pull is
straightforward.

We next claim\max{claim} that the relations $\inat$, $\itimes$, and $\iarr$ are
quasi-left-representable, and that their lifts are quasi-right-representable.
Indeed, it is straightforward to see that they are quasi-left-representable
where all of the perturbations are taken to be the identity elements of the
respective monoids.
%
For quasi-right-representability, the only nontrivial case is $\li(\iarr)$.
Recall the defintion of the downcast for $\iarr$ given in Section
\ref{sec:term-interpretation}.\max{why is there a paragraph break here?}

We know that in order for the DnR and DnL squares for $\li \iarr$ to exist, we
must insert a delay on the side opposite the downcast. In terms of syntactic
perturbations, this means that we take $\dellp$ and $\delrp$ in the definition
of quasi-right-representability to both be $\inl\, 1$, where $\inl$ is the
injection into the coproduct of monoids. We recall that the definition of
syntactic perturbations for $\li$ involves a coproduct with $\mathbb{N}$, and
that the interpretation homomorphism maps the generator $1$ to the endomorphism
of error domains $(\delta \circ \eta)^\dagger$.
%
With this choice for the perturbations, it is straightforward to show that the
DnL and DnR squares exist using the definition of the downcast and the
definition of the relation $\iarr$.

It is also straightforward to establish the retraction property for each of
these three relations. In the case of $\iarr$, we have that the property holds
up to a delay.\max{retraction property? has this ever come up anywhere in the paper before? Should it?}

@@

\eric{TODO: Does this belong here?}

Recall the equivalence rules for type precision given in Section \ref{sec:GTLC},
for example the rule that $(c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra
c_oc_o')$ To validate this rule in the model, we construct two squares $\id
\ltsqbisim{(c_i \ra c_o)(c_i' \ra c_o')}{(c_ic_i' \ra c_oc_o')} \id$ and $\id
\ltsqbisim{(c_ic_i' \ra c_oc_o')}{(c_i \ra c_o)(c_i' \ra c_o')} \id$. Since the
definition of extensional square allows us to specify intermediate morphisms
that are synchronized in the intensional ordering, it will be sufficient to
construct two intensional squares between the relations, with perturbations on
both sides. This motivates the definition of \emph{quasi-order-equivalence}.

\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
  between the underlying predomains. We say that $c$ and $c'$ are
  \textbf{quasi-order-equivalent}, written $c \qordeq c'$, if there exist
  syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
  \delta^r_2 \in M_{A'}$ such that there is a square $\delta^l_1 \ltsq{c}{c'}
  \delta^r_1$ and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
\end{definition}
Given computation objects $B$ and $B$, we make the analogous definition for
relations $d, d' : B \rel B'$ between the underlying error domains.

Then we have the following result that shows that the functors
$U,\li,\times,\to$ are \emph{quasi-functorial} on relations:

\begin{lemma}
The following hold:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
  \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
\end{itemize}

\end{lemma}
\begin{proof} See appendix \eric{TODO: make sure this is in the appendix}
\end{proof}

@@

In this section, we prove the adequacy of the interpretation of term precision.
First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong
\li \mathbb{N}$. We write $f \da n$ to mean that there exists $m$ such that $f =
\delta^m(\eta n)$ and $f \da \mho$ to mean that there exists $m$ such that $f =
\delta^m(\mho)$.

Recall that $\ltbisim$ denotes the relation on value morphisms defined as the
closure of the intensional error-ordering on morphisms under weak bisimilarity
on bboth sides. Specialized to the setting of morphisms from $1 \to U \li
\mathbb{N}$, we have $f \ltbisim g$ iff there exist $f'$ and $g'$ with
%
\[ f \bisim_{\li \mathbb{N}} f' \le_{\li \mathbb{N}} g' \bisim_{\li \mathbb{N}} g. \]\max{$\le$? I thought it was $\ltdyn$}
%

>>>>

We also enhance our error domains with ordering and bisimilarity
relation, and we impose a condition relating the $\theta$ morphism to
the bisimilarity relation. More concretely:
%
\begin{definition}
An \textbf{error domain} $B$ consists of a predomain $B$ along with the following data:
\begin{itemize}
    \item A distinguished ``error'' element $\mho_B \in B$ such that $\mho \le_B x$ for all $x$
    \item A morphism of predomains $\theta_B \colon \later B \to B$
    \item For all $x : B$ we have $\theta_B(\nxt\, x) \bisim_B x$
\end{itemize}\max{Don't we also need $\forall x. \mho \le x$?}
\end{definition}
%
For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B
\circ \nxt$.
%
A \textbf{morphism of error domains} is simply a morphism of the underlying
predomains that preserves the error element and the $\theta$ map.

@@

Next we define an analogous notion of \emph{relation} between error domains $B$
and $B'$. An error domain relation $d : B \rel B'$ is a relation on the
underlying posets that preserves error and respects $\theta$. More formally:
%
\begin{definition}
  Let $B$ and $B'$ be error domains. An \textbf{error domain relation} between
  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
  \begin{enumerate}
     \item (Respects error): For all $y \in B'$, we have $\mho_B \binrel{d} y$.\max{can we please use infix notation consistently for relations?}
     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t (\tilde{x}_t \binrel{d} \tilde{y}_t)$, then 
     $(\theta_B (\tilde{x}) \binrel{d} \theta_{B'} (\tilde{y}))$. 
  \end{enumerate}
\end{definition}
%
As with predomain relations, we can compose error domain relations. However, the
composition of the underlying predomain relations is not an error domain
relation.
% Can explain this further in the appendix
Instead, we define composition of error domain relations $d$ on $B_1$ and $B_2$
and $d'$ on $B_2$ and $B_3$ inductively to be the least relation containing $d$
and $d'$ that is downward closed under $\le_{B_1}$, upward-closed under
$\le_{B_3}$, respects error, and preserves $\theta$. See Definition
\ref{def:free-comp-ed-rel} in the appendix for the precise definition.

@@

The functors $U$, $\li$, $\times$, and $\arr$ introduced in the previous section
act on predomains and error domains and also have actions on their morphisms,
relations, and squares. The functor $U$ simply takes the underlying predomain
relation of an error domain relation. Given a relation $c_1 : A_1 \rel A_1'$ and
$c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2) \rel
(A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$ if
and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation respects error\max{preserves error?}
and preserves $\theta$ follows from the corresponding properties for $d$.

@@

To accomplish this, we equip every predomain $A$ with a monoid $M_A$ of
\emph{syntactic perturbations}, as well as a means of \emph{interpreting} these
syntactic perturbations as \emph{semantic} perturbations on $A$, i.e., as
endomorphisms that are bisimilar to the identity.
% Moreover, we want the
% resulting endomorphisms to be extensionally equivalent to the identity function,
% a condition we express by saying that they are weakly bisimilar to the identity
% morphism on $A$. 
More formally:
\begin{definition}[value and computation objects + morphisms]
  A \textbf{value object} consists of a predomain $A$, a monoid $M_A$ and a
  homomorphism of monoids $M_A \to \morbisimid{A} := \{ f : A \to A \mid f
  \bisim \id_A \}$.

  Likewise, a \textbf{computation object} consists of an error domain $B$, a
  monoid $M_B$ and a homomorphism of monoids $M_B \to \morbisimid{B}$.

  We define a \textbf{value morphism} to be a morphism of the underlying
  predomains, and a \textbf{computation morphism} to be a morphism of the
  underlying error domains.
\end{definition}

% every error domain $B$ with a monoid $M_B$ and interpretation homomorphism $i_B
% : M_B \to \morbisimid{B}$. This grouping of data deserves its own definition: We
% call the triple $(A, M_A, iA)$ a \textbf{value object} as it will be the
% denotation of value types in our final model. Likewise, the triple $(B, M_B,
% i_B)$ will be called a \textbf{computation object}.\max{make this a proper \textit{Definition} and define morphisms as well}

Having defined syntactic perturbations and value and computation objects, we can
now make formal the final aspect of the model: the notion of
quasi-represntability of relations.
%
\begin{definition}[quasi-left-representable relations]\label{def:quasi-left-representable}
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: (1) $\upl$: $e \ltsq{c}{\le_{A'}} i_{A'}(\delre_c)$, and
(2) $\upr$: $i_A(\delle_c) \ltsq{\le_{A}}{c} e$.
\end{definition}
%
Observe that this weakens the previous notion of representability, since under
that definition the perturbations were required to be the identity.
%
We make the analogous definition of quasi-left-representability for error domain
relations $d$.

@@

% \max{idea: move this after the quasi-representability part and say that this structure is what's needed to make them compose}
Given two quasi-representable relations $c$ and $c'$, it is not the case in
general that $cc'$ is quasi-representable. We need an additional condition that
specifies how relations interact with perturbations. In particular, we need to
be able to ``push'' and ``pull'' perturbations along relations $c$ and $d$.
%
The intuition for this requirement comes from the construction of the square
corresponding to the original UpL rule from the square for the simplified
version as was shown in Section \ref{sec:towards-relational-model}. Recall that
we horizontally composed the simplified UpL square for $c$ with the identity
square for $c'$ on the right. Now that the square for UpL involves a
perturbation on the right instead of the identity, this construction will not
work unless we can ``push'' the perturbation on $A_2$ to one on $A_3$. That is,
given a relation $c : A \rel A'$ and a syntactic perturbation $m_A \in M_A$, we
need to be able to turn it into a syntactic perturbation $m_{A'} \in M_{A'}$
such that the resulting semantic perturbations obtained by applying the
respective homomorphisms $i_A$ and $i_{A'}$ form a square. Dually, we will need
to ``pull'' a perturbation from right to left. This notion is made formal by the
following definition:
%
\begin{definition}
    Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects,
    and let $c : A \rel A'$ be a predomain relation. A
    \textbf{push-pull structure} for $c$ (with respect to $M_A$ and
    $M_{A'}$) consists of monoid homomorphisms $\push : M_A \to M_A'$
    and $\pull : M_A' \to M_A$ such that for all $m_A \in M_A$ there
    is a square $i_A(m_A) \ltsq{c}{c} i_{A'}(\push\, m_A)$ and for all
    $m_{A'} \in M_{A'}$ there is a square $i_A(\pull\, m_{A'})
    \ltsq{c}{c} i_{A'}(m_{A'})$. A push-pull structure for error
    domain relations is defined similarly.
\end{definition}

@@

% Because we have augmented the definitions of object and relation for our model,
% we need to specify the actions of the functors $F$, $U$, $\times$, and $\arr$ on
% these objects and relations. 
% For instance, the functor $F$ now acts on \emph{value objects} $(A, M_A, i_A)$,
% which means we need to specify the monoid $M_{FA}$ and interpretation $i_{FA}$
% corresponding to the error domain $FA$.

\subsubsection{Constructions involving Perturbations}\label{sec:perturbation-constructions}
\max{this section needs prose}
In this section, we discuss the actions of the functors on value and computation
objects. Specifically, we define the actions of the functors on the syntactic
perturbations and their interpretation as semantic perturbations.

We first discuss the action of $\li$ on objects. Given a value object $(A, M_A,
i_A)$ we define the syntactic perturbations $M_{\li A}$ to be $\mathbb{N} \oplus
M_A$, where $\oplus$ denotes the free product of monoids (the coproduct in the
category of monoids). The intuition behind needing to include $\mathbb{N}$ comes
from the example of the downcast for $\iarr$. Recall that we adjusted the
squares corresponding to the DnL and DnR rules by adding $(\delta \circ
\eta)^\dagger$ on the side opposite the downcast. Thus, we need a syntactic
perturbation in $M_{\li A}$ that will be interpreted as $(\delta \circ
\eta)^\dagger$. We accomplish this by taking the coproduct with $\mathbb{N}$.
%
Then to interpret the perturbations as endomorphisms, by the universal property
of the coproduct of monoids it suffices to define two homomorphisms, one from
$\mathbb{N} \to \morbisimid{\li A}$ and one from $M_A \to \morbisimid{\li A}$.
The first homomorphism sends the generator $1 \in \mathbb{N}$ to the error
domain endomorpism $(\delta \circ \eta)^\dagger$, and the second sends $m_A \in
M_A$ to $\li(i_A(m_A))$ and we observe in both cases that the resulting
endomorphisms are bisimilar to the identity since the action of $\li$ on
morphisms preserves bisimilarity.

@@

The fact that value and computation relations compose is proved in the appendix
(Definition \ref{def:value-computation-rel-comp}). The actions of the functors
$\li$, $U$, $\times$, and $\arr$ are also defined in the appendix (Definition
\ref{def:functorial-actions-on-relations}).

\subsection{The Dynamic Type}
We now describe the denotation of the dynamic type. Analogously to how we
defined the dynamic type in Section \ref{sec:concrete-term-model}, the predomain
representing the dynamic type is defined using a combination of least fixpoint
and guarded recursion as the solution to the equation of \emph{predomains}
%
\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr \li D). \]
%
As before, we name the constructors $\inat$, $\itimes$, and $\iarr$
respectively, and the upcasts $e_{\nat}$, $e_\times$, and $e_\to$ are the same
as the constructors except for $e_{\iarr}$ which includes a $\nxt$.
%
%
% We define $e_\mathbb{N} : \mathbb{N} \to D$ to be the $\text{nat}$ constructor,
% $e_\times : D \times D \to D$ to be $\text{times}$, and $e_\to : U(D \arr F D)$
% to be the morphism $\nxt$ followed by $\text{fun}$.
%
The definition of the ordering $\le_D$ is provided in the appendix for completeness.

% given by:\max{this is kind of obvious IMO maybe it can go in the appendix}
%

%
%
We define three relations involving $D$ as follows. We define $\inat :
\mathbb{N} \rel D$ by $(n, d) \in \inat$ iff $e_\nat(n) \le_D d$. We similarly
define $\itimes : D \times D \rel D$ by $((d_1, d_2), d) \in \itimes$ iff
$e_\times(d_1, d_2) \le_D d$, and we define $\text{inj}_\to : U(D \arr \li D)
\rel D$ by $(f, d) \in \iarr$ iff $e_\to(f) \le_D d$.

Now we define the perturbations for $D$. Recall that to each predomain $A$ we
associate a monoid $M_A$ of syntactic perturbations and a homomorphism into the
monoid of endomorphisms bisimilar to the identity, and likewise for error
domains. It may seem as though we need to define the perturbations for $D$ by
guarded recursion, but in fact we define them via least-fixpoint in the category
of monoids:
%
\( M_D \cong (M_{D \times D}) \oplus M_{U(D \to \li D)}. \)
%
We now explain how to interpret these perturbations as endomorphisms.

@@

One can verify that this defines a homomorphism from $M_D \to \{ f : D \to D : f
\bisim \id \}$. Then it is straightforward to check that the three relations $\inat$, $\itimes$, and
$\iarr$ satisfy the push-pull property. As an illustrative case, we establish
the push-pull property for the relation $\iarr$. We define $\pull_{\iarr} : M_D
\to M_{U(D \arr \li D)}$ by giving a homomorphism from $M_{D \times D} \to
M_{U(D \arr \li D)}$ and from $M_{U(D \to \li D)} \to M_{U(D \arr \li D)}$. The
former is the trivial homomorphism sending everything to the identity element,
while the latter is the identity homomorphism.
%
Conversely, we define $\push_{\iarr} : M_{U(D \arr \li D)} \to M_D$ as the
inclusion into the coproduct.
%
Showing that the relevant squares exist involving push and pull is
straightforward.

It is easy to see that the relations $\inat$, $\itimes$, and $\iarr$
are quasi-left-representable, where all perturbations are taken to be the
identity elements of the respective monoids.
%
For quasi-right-representability, the only nontrivial case is $\li(\iarr)$.
Recall the defintion of the downcast for $\iarr$ given in Section
\ref{sec:term-interpretation}. We know that in order for the DnR and DnL squares
for $\li \iarr$ to exist, we must insert a delay on the side opposite the
downcast. In terms of syntactic perturbations, this means that we take $\dellp$
and $\delrp$ in the definition of quasi-right-representability to both be
$\inl\, 1$, where $\inl$ is the left injection into the coproduct of monoids. We
recall that the definition of syntactic perturbations for $\li$ involves a
coproduct with $\mathbb{N}$, and that the interpretation homomorphism maps the
generator $1$ to the endomorphism of error domains $(\delta \circ
\eta)^\dagger$.
%
With this choice for the perturbations, it is straightforward to show that the
DnL and DnR squares exist using the definition of the downcast and the
definition of the relation $\iarr$.

@@

\max{Missing a subsection on how to interpret the syntax into the semantics, i.e., the actual graduality proof}
\subsection{Interpreting the Relational Axioms}

We now turn to the interpretation of the type precision and term precision
axioms of the cast calculus (Section \ref{sec:GTLC}).

The congruence rules for term precision hold in the model, as all of
constructions are monotone and preserve weak bisimilarity.
%
The interpretation of the cast rules was discussed above.
%

It remains to verify the equivalence rules for type precision. For example,
consider the rule $(c_i \ra c_o)(c_i' \ra c_o')\equiv (c_ic_i' \ra c_oc_o')$. To
validate this rule in the model, we construct two squares $\id \ltsqbisim{(c_i
\ra c_o)(c_i' \ra c_o')}{(c_ic_i' \ra c_oc_o')} \id$ and $\id
\ltsqbisim{(c_ic_i' \ra c_oc_o')}{(c_i \ra c_o)(c_i' \ra c_o')} \id$. Since the
definition of extensional square allows us to specify intermediate morphisms
that are synchronized in the intensional ordering, it will be sufficient to
construct two intensional squares between the relations, with perturbations on
both sides. This motivates the definition of \emph{quasi-order-equivalence}.

\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
  between the underlying predomains. We say that $c$ and $c'$ are
  \textbf{quasi-order-equivalent}, written $c \qordeq c'$, if there exist
  syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
  \delta^r_2 \in M_{A'}$ such that there is a square $i_A(\delta^l_1)
  \ltsq{c}{c'} i_{A'}(\delta^r_1)$ and a square $i_A(\delta^l_2) \ltsq{c'}{c}
  i_{A'}(\delta^r_2)$. Given computation objects $B$ and $B$, we make the
  analogous definition for relations $d, d' : B \rel B'$ between the underlying
  error domains.
\end{definition}

Then we have the following quasi-order-equivalences:
%
\begin{lemma}
The following hold:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
  \item $\li(c \comp c') \qordeq \li (c) \comp \li (c')$
  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
\end{itemize}
\end{lemma}
\begin{proof}
  See appendix (Lemma \ref{lem:quasi-order-equiv-functors}).
\end{proof}

@@

In this section, we prove the adequacy of the interpretation of term precision.

\begin{comment}
First we establish some notation. Fix a morphism $f : 1 \to \li \mathbb{N} \cong
\li \mathbb{N}$. We write $f \da n$ to mean that there exists $m$ such that $f =
\delta^m(\eta n)$ and $f \da \mho$ to mean that there exists $m$ such that $f =
\delta^m(\mho)$.
